---
title: "Patient2_analysis"
author: "bhaas"
date: '2023-09-28'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


```{r}
MIN_CELLS = 5
```


# P2 Tum

Only a tumor sample for P2, no matching normal

```{r}

# parse fusion evidence support including cell barcode and umi 

Tum_data = read.table("data/Patient2_Tum.Dondi_overian_CTAT_fusions.filtered_cells_and_dedup_umis.tsv.gz", header=T, sep="\t", stringsAsFactors = F) %>%
    filter(! is.na(UMAP_1))

nrow(Tum_data)

head(Tum_data)


```

```{r}
# since starF and FI were run at max sensitivity, lets restrict fusions to those identified by ctat-LRF

Tum_ctat_LRF_fusion_genes = Tum_data %>% filter(method == 'ctat-LR-fusion') %>% select(FusionName) %>% unique() %>% pull(FusionName)

Tum_data = Tum_data %>% filter(FusionName %in% Tum_ctat_LRF_fusion_genes)

nrow(Tum_data)

```


```{r}

fusion_annots = read.csv("data/Patient2.fusion_annots.gz", sep="\t", header=T, stringsAsFactors = F)

```



```{r}

# parse UMAP info (from Dondi et al.) 

Tum_umap_data = read.table("data/Patient2_Tum_UMAPcoords.txt.gz", header=T, sep=",") %>%
    rename(barcodes=X)
celltypes = read.table("data/Patient2_Tum.bc_to_celltype.tsv.gz", header=T, sep="\t")

Tum_umap_data = full_join(Tum_umap_data, celltypes, by='barcodes')


# number of cells
num_tumor_cells = nrow(Tum_umap_data) 
message("number tumor cells: ", num_tumor_cells)
# 453 total cells

Tum_umap_data %>% group_by(celltype_final) %>% tally(name='count_cell_type') %>% mutate(frac_tot_cells = count_cell_type/num_tumor_cells)

# 208 HGSOC cells  (46% of total cells)

```


```{r}

Tum_cell_counts = Tum_data %>% select(FusionName, cell_barcode) %>% unique() %>% 
    group_by(FusionName) %>%
    tally(name='tot_cells_w_fusion') %>%
    mutate(frac_tot_cells=tot_cells_w_fusion/num_tumor_cells)  %>%
    arrange(desc(frac_tot_cells))

left_join(Tum_cell_counts %>% filter(tot_cells_w_fusion >= MIN_CELLS),
          fusion_annots,
          by='FusionName')
             
```


```{r}

# examine distribution of fusion calls according to cell types

Tum_fusion_frac_cell_types = Tum_data %>% select(FusionName, barcodes, celltype_final) %>% unique() %>%
    group_by(FusionName, celltype_final) %>% tally(name='tot_cells_w_fusion') %>% 
    mutate(frac_fusion_cells=prop.table(tot_cells_w_fusion)) %>%
    arrange(desc(tot_cells_w_fusion))

Tum_fusion_frac_cell_types %>% head()


```



```{r}

Tum_data %>% select(method) %>% unique()

```

```{r}

starF_fusions = Tum_data %>% filter(method=="STAR-Fusion")

FI_fusions = Tum_data %>% filter(method=="FusionInspector")

ctat_LRF_fusions = Tum_data %>% filter(method == "ctat-LR-fusion")


```




```{r}

# counts of cells according to fusion breakpoint and method for detection

Tum_cell_counts_by_method = read.table("data/Patient2_Tum.Dondi_overian_CTAT_fusions.filtered_cells_and_dedup_umis.cell_counts_by_method.tsv.gz",
                                          header=T, sep="\t", stringsAsFactors = F)


Tum_cell_counts_by_method %>% head()

```

```{r}

# reorganizing cell counts of fusions by method for comparison

Tum_cell_counts_by_method_spread  = Tum_cell_counts_by_method %>% select(FusionName, LeftBreakpoint, RightBreakpoint, method, celltype_final, cell_counts) %>%
    spread(key=method, value=cell_counts) %>% 
    arrange(desc(`ctat-LR-fusion`)) 


Tum_cell_counts_by_method_spread  %>% filter(`ctat-LR-fusion` >= MIN_CELLS)

```



# plot counts of cells for these fusions:

```{r}

# examine counts of cells according to fusion and method lmited to those with at least the min number of cells

right_join(Tum_cell_counts_by_method, 
          Tum_cell_counts_by_method %>% 
                         filter(cell_counts >= MIN_CELLS)  %>% 
                         select(FusionName, LeftBreakpoint, RightBreakpoint) 
          ) %>%
              rowwise() %>% mutate(fusion=paste(FusionName, LeftBreakpoint, RightBreakpoint, collapse=":")) %>%
              ggplot(aes(x=fusion, y=cell_counts, fill=method)) + geom_bar(stat='identity', position='dodge') +
              theme(axis.text.x = element_text(angle = 90, hjust = 1))


```



# Examine cell type representation by fusions


```{r}

# counts of cells containing fusions and fraction of fusions identified according to that cell type


fusion_frac_cell_types = Tum_data %>% select(FusionName, barcodes, celltype_final) %>% unique() %>%
    group_by(FusionName, celltype_final) %>% tally(name='tot_cells_w_fusion') %>% 
    mutate(frac_fusion_cells=prop.table(tot_cells_w_fusion)) %>%
    arrange(desc(tot_cells_w_fusion))

fusion_frac_cell_types %>% filter(tot_cells_w_fusion >= MIN_CELLS)

```




```{r}

# restrict to fusions of interest:  those showing up in at least min cells and at least 80% representation within any cell type

fusions_of_interest = fusion_frac_cell_types %>% 
    filter(tot_cells_w_fusion >= MIN_CELLS) %>%
    arrange(desc(tot_cells_w_fusion)) %>%
    filter(frac_fusion_cells >= 0.8)

fusions_of_interest

# 17 fusions of interest

```


```{r}

# include fusion annotations

fusions_of_interest = left_join(fusions_of_interest,
          fusion_annots)

fusions_of_interest
```

```{r}

write.table(fusions_of_interest, file="Patient2_Tum.fusions_of_interest.tsv", sep="\t", quote=F, row.names = F)


```






```{r}

# examine method of detection and cell counts for just the fusions of interest

Tum_cell_counts_by_method %>% 
    filter(FusionName %in% fusions_of_interest$FusionName) %>%
    select(FusionName, LeftBreakpoint, RightBreakpoint, method, celltype_final, cell_counts) %>%
    spread(key=method, value=cell_counts) %>% 
    arrange(desc(`ctat-LR-fusion`))


```



```{r}

# plotting counts of cells according to method for fusions of interest

Tum_cell_counts_by_method %>% 
            filter(FusionName %in% fusions_of_interest$FusionName) %>%
            select(FusionName, LeftBreakpoint, RightBreakpoint, cell_counts, method) %>%
              rowwise() %>% mutate(fusion=paste(FusionName, LeftBreakpoint, RightBreakpoint, collapse=":")) %>%
              ggplot(aes(x=fusion, y=cell_counts, fill=method)) + 
                 geom_bar(stat='identity', position='dodge') +
                 theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Examine fusions of interest in UMAPs





```{r}
# base umap plot coloring by annotated cell type

baseplot = Tum_umap_data %>% ggplot(aes(x=UMAP_1, y=UMAP_2)) + geom_point(aes(color=celltype_final))

baseplot



```


```{r}

# Examine each fusion of interest according to UMAP cell positioning

x = 0

plots = list()

for (fusion in  fusions_of_interest$FusionName) {
    
    p = baseplot + geom_point(data=Tum_data %>% filter(FusionName == fusion) %>% select(UMAP_1, UMAP_2) %>% unique(), 
                              color='red') + 
        ggtitle(paste("Patient 2 Tumor, Fusion: ", fusion) )
    
    plot(p)   
    
    x = x+1
    plots[[x]] = p
}



```


```{r}

# make a pdf containing these plots

pdf("Patient2_Tum.fusions_of_interest.pdf")
for (p in plots) {
    plot(p)
}

dev.off()

```





```{r}

# examine counts of cells by method for the fusions of interest

tumor_cell_counts = Tum_data %>% filter(FusionName %in%  fusions_of_interest$FusionName) %>% 
        select(FusionName, method, barcodes) %>% unique() %>%
        group_by(FusionName, method) %>% tally(name='cell_counts')


tumor_cell_counts %>% spread(key=method, value=cell_counts) %>% arrange(desc(`ctat-LR-fusion`))

```


```{r}

# plot counts of cells by method for fusions of interest

tumor_cell_counts %>%
              ggplot(aes(x=FusionName, y=cell_counts, fill=method)) + geom_bar(stat='identity', position='dodge') +
              theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle("Patient2_Tum Fusions of Interest: Cell Counts")


```


```{r}

# Examine Venn for cells detected according to combinations of methods.

tumor_cell_counts_by_methods = Tum_data %>% filter(FusionName %in%  fusions_of_interest$FusionName) %>% 
        select(FusionName, method, barcodes) %>% unique() %>%
        group_by(FusionName, barcodes) %>% 
        arrange(method) %>%
        mutate(methods = paste(method, collapse=',')) %>%
        ungroup() %>%
        select(FusionName, methods, barcodes) %>% group_by(FusionName, methods) %>% tally()


tumor_cell_counts_by_methods 

```


# Examine specific fusions of interest

## IGF2BP2--TESPA1

```{r}

tumor_cell_counts_by_methods %>% filter(FusionName == "IGF2BP2--TESPA1")

```
```{r}

Tum_cell_counts_by_method %>% 
    filter(FusionName=="IGF2BP2--TESPA1") %>%
    select(FusionName, LeftBreakpoint, RightBreakpoint, method, celltype_final, cell_counts) %>%
    spread(key=method, value=cell_counts) %>% 
    arrange(desc(`ctat-LR-fusion`), desc(FusionInspector))

```

## SPATS2--TRA2B


```{r}

tumor_cell_counts_by_methods %>% filter(FusionName == "SPATS2--TRA2B")


```


```{r}


Tum_cell_counts_by_method %>% 
    filter(FusionName=="SPATS2--TRA2B") %>%
    select(FusionName, LeftBreakpoint, RightBreakpoint, method, celltype_final, cell_counts) %>%
    spread(key=method, value=cell_counts) %>% 
    arrange(desc(`ctat-LR-fusion`), desc(FusionInspector))


```



## Are SPATS2--TRA2B and IGF2BP2--TESPA1 found expressed in the same tumor cells?

```{r}

Tum_data %>% filter(FusionName %in% c('SPATS2--TRA2B', 'IGF2BP2--TESPA1')) %>%
    select(FusionName, cell_barcode) %>%
    unique() %>%
    group_by(cell_barcode) %>% 
    arrange(FusionName) %>%
    mutate(fusion_names = paste(FusionName, collapse=",")) %>%
    ungroup() %>%
    select(cell_barcode, fusion_names) %>% unique() %>%
    group_by(fusion_names) %>% tally()
    



```

# Are there interesting fusions in non-tumor cells?

```{r}

fusion_frac_cell_types %>% filter(celltype_final != "HGSOC") %>% arrange(desc(tot_cells_w_fusion)) %>% head()
 
```

No fusions showing up in non-HGSOC cells at the min cell cutoff.  The top two have 4 cells and are found mostly in HGSOC.
Besides those, there's a max of 2 cells having any other fusion transcript.



```{r}

report_on_fusion = function(fusion_name) {
    
    print(Tum_cell_counts_by_method_spread %>% filter(FusionName == fusion_name))
    
    print(Tum_cell_counts %>% filter(FusionName == fusion_name) %>% mutate(type="Tum"))
    
    print(Tum_fusion_frac_cell_types %>% filter(FusionName == fusion_name) %>% mutate(type="Tum"))
    
}

```




# queries on specific fusions found in normal samples


```{r}
report_on_fusion("RP11-444D3.1--SOX5")

```

```{r}

report_on_fusion("RP11-208G20.2--PSPHP1")

```

```{r}

report_on_fusion("PSMB7--SCAI")

```




