---
title: "Patient1_analysis"
author: "bhaas"
date: '2023-09-28'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```



```{r}
MIN_CELLS = 5
```



# P1 Om


```{r}

# Om sample fusion read evidence with cell barcodes and UMIs

Om_data = read.table("data/Patient1_Om.Dondi_overian_CTAT_fusions.filtered_cells_and_dedup_umis.tsv.gz", header=T, sep="\t", stringsAsFactors = F) %>% 
    filter(! is.na(UMAP_1))

nrow(Om_data)

Om_data %>% head()


```

```{r}
# since starF and FI were run at max sensitivity, lets restrict fusions to those identified by ctat-LRF

Om_ctat_LRF_fusion_genes = Om_data %>% filter(method == 'ctat-LR-fusion') %>% select(FusionName) %>% unique() %>% pull(FusionName)

Om_data = Om_data %>% filter(FusionName %in% Om_ctat_LRF_fusion_genes)

nrow(Om_data)

```

```{r}

fusion_annots = read.table("data/Patient1.fusion_annots.gz", header=T, sep="\t", stringsAsFactors = F)

```




```{r}
# Om sample composition

Om_bc_to_celltype = read.table("data/Patient1_Om.bc_to_celltype.tsv.gz", header=T, sep="\t")
Om_umap_coords = read.table("data/Patient1_Om_UMAPcoords.txt.gz", header=T, sep=",") %>% rename(barcodes=X)

Om_umap_data = left_join(Om_umap_coords, Om_bc_to_celltype, by='barcodes')

num_Om_cells = nrow(Om_umap_data)

message("Have ", num_Om_cells, " Om cells")
# 594 Om cells

Om_data = Om_data %>% filter(barcodes %in% Om_umap_data$barcodes)
nrow(Om_data)
```

```{r}

Om_cell_counts = Om_data %>% select(FusionName, cell_barcode) %>% unique() %>% 
    group_by(FusionName) %>%
    tally(name='tot_cells_w_fusion') %>%
    mutate(frac_tot_cells=tot_cells_w_fusion/num_Om_cells) %>%
    arrange(desc(frac_tot_cells))

Om_cell_counts %>% filter(tot_cells_w_fusion >= MIN_CELLS)
```




```{r}

# Om sample counts of cells with fusions according to fusion detection method.

Om_cell_counts_by_method = read.table("data/Patient1_Om.Dondi_overian_CTAT_fusions.filtered_cells_and_dedup_umis.cell_counts_by_method.tsv.gz", 
                                      header=T, sep="\t", stringsAsFactors = F)

Om_cell_counts_by_method %>% head()

```


```{r}

# reorganizing cell counts by method for comparing across methods and corresponding types of reads (long or short)

Om_cell_counts_by_method_spread = Om_cell_counts_by_method %>% select(FusionName, LeftBreakpoint, RightBreakpoint, method, celltype_final, cell_counts) %>%
    spread(key=method, value=cell_counts) %>% 
    arrange(desc(`ctat-LR-fusion`))

Om_cell_counts_by_method_spread %>% filter(`ctat-LR-fusion` >= MIN_CELLS)
```

```{r}

# determine how each fusion is distributed across different cell types.

Om_fusion_frac_cell_types = Om_data %>% select(FusionName, barcodes, celltype_final) %>% unique() %>%
    group_by(FusionName, celltype_final) %>% tally(name='tot_cells_w_fusion') %>% 
    mutate(frac_fusion_cells=prop.table(tot_cells_w_fusion)) %>%
    arrange(desc(tot_cells_w_fusion))

Om_fusion_frac_cell_types %>% filter(tot_cells_w_fusion >= MIN_CELLS) %>% arrange(FusionName, desc(tot_cells_w_fusion))


```



# P1 Tumor


```{r}

# read in tumor fusion read evidence with cell barcode and UMIs


Tum_data = read.table("data/Patient1_Tum.Dondi_overian_CTAT_fusions.filtered_cells_and_dedup_umis.tsv.gz", header=T, sep="\t", stringsAsFactors = F) %>%
    filter(! is.na(UMAP_1))

nrow(Tum_data)

Tum_data %>% head()

```

```{r}
# since starF and FI were run at max sensitivity, lets restrict fusions to those identified by ctat-LRF

Tum_ctat_LRF_fusion_genes = Tum_data %>% filter(method == 'ctat-LR-fusion') %>% select(FusionName) %>% unique() %>% pull(FusionName)

Tum_data = Tum_data %>% filter(FusionName %in% Tum_ctat_LRF_fusion_genes)

nrow(Tum_data)

```



```{r}

Tum_umap_data = read.table("data/Patient1_Tum_UMAPcoords.txt.gz", header=T, sep=",") %>%
    rename(barcodes=X) %>%
    filter(! is.na(UMAP_1))

celltypes = read.table("data/Patient1_Tum.bc_to_celltype.tsv.gz", header=T, sep="\t")

Tum_umap_data = left_join(Tum_umap_data, celltypes, by='barcodes')

# number of cells
num_tumor_cells = nrow(Tum_umap_data) 
message("number total cells: ", num_tumor_cells)
# 497 total cells

Tum_umap_data %>% group_by(celltype_final) %>% tally(name='count_cell_type') %>% mutate(frac_tot_cells = count_cell_type/num_tumor_cells)

# 92 HGSOC cells  (19% of total cells)

```


```{r}

Tum_cell_counts = Tum_data %>% select(FusionName, cell_barcode) %>% unique() %>% 
    group_by(FusionName) %>%
    tally(name='tot_cells_w_fusion') %>%
    mutate(frac_tot_cells=tot_cells_w_fusion/num_tumor_cells)  %>%
    arrange(desc(frac_tot_cells))

left_join(Tum_cell_counts %>% filter(tot_cells_w_fusion >= MIN_CELLS),
          fusion_annots,
          by='FusionName')
```


```{r}

# read in cell counts by method

Tum_cell_counts_by_method = read.table("data/Patient1_Tum.Dondi_overian_CTAT_fusions.filtered_cells_and_dedup_umis.cell_counts_by_method.tsv.gz", 
                                      header=T, sep="\t", stringsAsFactors = F)

Tum_cell_counts_by_method %>% head()

```

```{r}

# organize fusion cell counts by method to compare methods


Tum_cell_counts_by_method_spread = Tum_cell_counts_by_method %>% select(FusionName, LeftBreakpoint, RightBreakpoint, method, celltype_final, cell_counts) %>%
    spread(key=method, value=cell_counts) %>% 
    arrange(desc(`ctat-LR-fusion`))

Tum_cell_counts_by_method_spread %>% filter(`ctat-LR-fusion` >= MIN_CELLS)
```



# compare P1 Tum and Om fusions

```{r}

# join the Om and Tum fusion cell counts by method

Tum_n_Om_joined_fusions = full_join(Tum_cell_counts_by_method_spread, Om_cell_counts_by_method_spread, 
                                    by=c('FusionName', 'LeftBreakpoint', 'RightBreakpoint', 'celltype_final'),
                                    suffix=c('.Tum', '.Om'))

Tum_n_Om_joined_fusions %>% head()

```

```{r}

# examine those fusions with at least MIN_CELLS by long reads in the tumor sample

Tum_n_Om_joined_fusions %>% select(FusionName, LeftBreakpoint, RightBreakpoint, celltype_final, `ctat-LR-fusion.Tum`, `ctat-LR-fusion.Om`) %>%
    arrange(desc(`ctat-LR-fusion.Tum`)) %>%
    filter(`ctat-LR-fusion.Tum` >= MIN_CELLS)


```


```{r}

# examine distribution of fusion calls according to cell types

Tum_fusion_frac_cell_types = Tum_data %>% select(FusionName, barcodes, celltype_final) %>% unique() %>%
    group_by(FusionName, celltype_final) %>% tally(name='tot_cells_w_fusion') %>% 
    mutate(frac_fusion_cells=prop.table(tot_cells_w_fusion)) %>%
    arrange(desc(tot_cells_w_fusion))

Tum_fusion_frac_cell_types %>% head()


```


```{r}
# identify tumor-enriched fusions:
# restrict to those where at least 80% of the fusion containing cells are tumor HGSOC cells


fusions_of_interest = Tum_fusion_frac_cell_types %>% filter(celltype_final == "HGSOC" & frac_fusion_cells >= 0.8) %>%
    arrange(desc(tot_cells_w_fusion)) %>%
    filter(tot_cells_w_fusion >= MIN_CELLS)



write.table(left_join(fusions_of_interest, fusion_annots), file="Patient1_Tum.fusions_of_interest.tsv", sep="\t", row.names=F, quote=F)


fusions_of_interest
```

only 3 fusions of interest, each with at least 5 cells, and found entirely within the tumor cell fraction

```{r}

# see if these fusions are found in the Om sample

left_join(fusions_of_interest, Om_fusion_frac_cell_types, by='FusionName', suffix=c('.Tum', '.Om'))


```

None of these three fusions are found in the matched normal 'Om' sample


# Examine read type support for these fusions of interest

```{r}

# see if we find these fusions using short reads:

fusions_of_interest = left_join(fusions_of_interest, Tum_cell_counts_by_method_spread)

fusions_of_interest = left_join(fusions_of_interest, fusion_annots)

fusions_of_interest
```


RAPGEF5--AGMO was found by short and long reads. The others were found only by the long reads.

# Examine umaps for fusions




```{r}
# add seurat clusters

seurat_clusters = read.table("data/Patient1_Tum.umap_cluster_ids.tsv.gz", header=T, sep="\t")

Tum_umap_data = full_join(Tum_umap_data, seurat_clusters, by='barcodes') %>% rename(cluster_id = seurat_clusters) 

```


```{r}
baseplot = Tum_umap_data %>% ggplot(aes(x=UMAP_1, y=UMAP_2)) + geom_point(aes(color=celltype_final))

baseplot

```


```{r}


x = 0

plots = list()

for (fusion in  fusions_of_interest$FusionName) {
    
    p = baseplot + geom_point(data=Tum_data %>% filter(FusionName == fusion) %>% select(UMAP_1, UMAP_2) %>% unique(), 
                              color='red') + 
        ggtitle(paste("Patient 1 Tumor, Fusion: ", fusion) )
    
    plot(p)   
    
    x = x+1
    plots[[x]] = p
}



```



```{r}

pdf("Patient1_Tum.fusions_of_interest.pdf")
for (p in plots) {
    plot(p)
}

dev.off()

```


```{r}

fusion_of_interest_cell_counts = Tum_data %>% filter(FusionName %in%  fusions_of_interest$FusionName) %>% 
        select(FusionName, method, barcodes) %>% unique() %>%
        group_by(FusionName, method) %>% tally(name='cell_counts') 


fusion_of_interest_cell_counts %>% spread(key=method, value=cell_counts) %>% arrange(desc(`ctat-LR-fusion`))




```

```{r}

fusion_of_interest_cell_counts  %>%
              ggplot(aes(x=FusionName, y=cell_counts, fill=method)) + geom_bar(stat='identity', position='dodge') +
              theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle("Patient1_Tum Fusions of Interest: Cell Counts")



```

```{r}

# breakdown by combinations of methods per cell

tumor_cell_counts_by_methods = Tum_data %>% filter(FusionName %in%  fusions_of_interest$FusionName) %>% 
        select(FusionName, method, barcodes) %>% unique() %>%
        group_by(FusionName, barcodes) %>% 
        arrange(method) %>%
        mutate(methods = paste(method, collapse=',')) %>%
        ungroup() %>%
        select(FusionName, methods, barcodes) %>% unique() %>% group_by(FusionName, methods) %>% tally()


tumor_cell_counts_by_methods 


```

For RAPGEF5::AGMO, half (3/6) of the cells were detected only by long reads, and 1/6 exclusively by short reads.
 
Fusions SMG7::CH507-513H4.1 and NTN1::CDRT15P2 were only detected by long reads.   
 

```{r}
p = baseplot +
        
    geom_point(data=Tum_data %>% filter(FusionName %in% c('SMG7--CH507-513H4.1', 'RAPGEF5--AGMO')) %>% select(FusionName, UMAP_1, UMAP_2) %>% unique(), 
                              aes(shape=FusionName), alpha=0.7) +
         ggtitle("Patient 1")


p
```

```{r}

ggsave(p, file="HGSOC_Patient1_umap.svg", width=7, height=5)


```



```{r}

tumor_cell_counts_by_methods  %>% filter(FusionName == "SMG7--CH507-513H4.1")

```





```{r}

tumor_cell_counts_by_methods  %>% filter(FusionName == "RAPGEF5--AGMO")

```


```{r}

tumor_cell_counts_by_methods  %>% filter(FusionName == "NTN1--CDRT15P2")

```


```{r}


baseplot = Tum_umap_data %>% ggplot(aes(x=UMAP_1, y=UMAP_2)) + geom_point(aes(color=factor(cluster_id)))


x = 0

#plots = list()

for (fusion in  fusions_of_interest$FusionName) {
    
    p = baseplot + geom_point(data=Tum_data %>% filter(FusionName == fusion) %>% select(UMAP_1, UMAP_2) %>% unique(), 
                              color='red') + 
        ggtitle(paste("Patient 1 Tumor, Fusion: ", fusion) )
    
    plot(p)   
    
    x = x+1
    #plots[[x]] = p
}


```


The SMG7 and RAPGEF5--AGMO fusions appear in different tumor subclusters.  

The NTN1 fusion is found across both subclusters.



# fusions found in normal cells?

```{r}

left_join(Tum_n_Om_joined_fusions %>% filter(`ctat-LR-fusion.Tum` >= MIN_CELLS & `ctat-LR-fusion.Om` > 0),
          fusion_annots) 


```

```{r}

left_join(Tum_n_Om_joined_fusions %>% filter(`ctat-LR-fusion.Tum` >= MIN_CELLS & `ctat-LR-fusion.Om` > 0),
          fusion_annots)  %>%
    select(FusionName, annots) %>% unique()

```


```{r}

# top celltype count only for each
left_join(Tum_n_Om_joined_fusions %>% filter(`ctat-LR-fusion.Tum` >= MIN_CELLS & `ctat-LR-fusion.Om` > 0),
          fusion_annots,
          by='FusionName')   %>%
    group_by(FusionName) %>% arrange(desc(`ctat-LR-fusion.Tum`)) %>% filter(row_number()==1) %>% ungroup() %>%
    
    select(FusionName, `ctat-LR-fusion.Tum`, celltype_final, annots) %>% unique()

```



```{r}

report_on_fusion = function(fusion_name) {
    
    print(left_join(Tum_n_Om_joined_fusions %>% filter(FusionName == fusion_name),
              fusion_annots, by='FusionName') )
    
    print(Om_cell_counts %>% filter(FusionName == fusion_name) %>% mutate(type="Om"))
    
    print(Om_fusion_frac_cell_types %>% filter(FusionName == fusion_name) %>% mutate(type="Om"))
    
    print(Tum_cell_counts %>% filter(FusionName == fusion_name) %>% mutate(type="Tum"))
    
    print(Tum_fusion_frac_cell_types %>% filter(FusionName == fusion_name) %>% mutate(type="Tum"))
    
}

```


## GS1-279B7.2--GNG4

```{r}

report_on_fusion("GS1-279B7.2--GNG4")

# GS1-279B7.2--GNG4	INTRACHROMOSOMAL[chr1:50.23Mb]

```



##  RP11-208G20.2--PSPHP1

```{r}

report_on_fusion("RP11-208G20.2--PSPHP1")


```


##  RP11-384F7.2--LSAMP

```{r}

report_on_fusion("RP11-384F7.2--LSAMP")

```


## RP1-34H18.1--NAV3

```{r}

report_on_fusion("RP1-34H18.1--NAV3")


```



# normal fusions in the matched Om sample?


```{r}

# Om normal sample fusions

left_join(Om_fusion_frac_cell_types %>% filter(tot_cells_w_fusion >= MIN_CELLS) %>% arrange(FusionName, desc(tot_cells_w_fusion)) %>%
    group_by(FusionName) %>% arrange(desc(tot_cells_w_fusion)) %>% filter(row_number()==1) %>% ungroup(),
    fusion_annots,
    by='FusionName')


```

YWHAE--CRK is interesting as found before among different tumor data sets and involves a pair of known oncogenes.



```{r}

report_on_fusion("YWHAE--CRK")

```

## RP11-444D3.1--SOX5 
```{r}

report_on_fusion("RP11-444D3.1--SOX5")

```



# Tumor heterogeneity explored

```{r}

tumor_umap = read.table("data/Patient1_Tum_only/Patient1_Tum_Cancer_only.csv", header=T, sep=",") %>%
    rename(barcodes=X)

tumor_umap %>% head()


```
```{r}

tumor_umap = left_join(tumor_umap, 
                       Tum_data, 
                       by='barcodes', 
                       suffix=c('.tum_only', '.allcells'))

tumor_umap %>% head()

```
```{r}

tumor_umap %>% select(barcodes, UMAP_1.allcells, UMAP_2.allcells) %>% unique() %>% 
    ggplot(aes(x=UMAP_1.allcells, y=UMAP_2.allcells)) +
    geom_point() + ggtitle("UMAP for tumor subset in all-cells umap")

```
```{r}


tum_only_umap_baseplot = tumor_umap %>% select(barcodes, UMAP_1.tum_only, UMAP_2.tum_only, seurat_clusters) %>% unique() %>% 
    ggplot(aes(x=UMAP_1.tum_only, y=UMAP_2.tum_only, shape=as.factor(seurat_clusters))) +
    geom_point(size=rel(2)) + ggtitle("UMAP for tumor subset in tum-only umap")


tum_only_umap_baseplot

```


```{r}


tumor_cells_only_fig = tum_only_umap_baseplot + 
    geom_point(data=tumor_umap %>% filter(FusionName %in% c('SMG7--CH507-513H4.1', 'RAPGEF5--AGMO', 'NTN1--CDRT15P2')) %>% 
                   select(FusionName, UMAP_1.tum_only, UMAP_2.tum_only, seurat_clusters) %>% unique(), 
                              aes(color=FusionName), size=rel(5), alpha=0.3) +
         ggtitle("Patient 1 Tumor Cells Only")

tumor_cells_only_fig
```


```{r}

ggsave(tumor_cells_only_fig, file="scHGSOC_Patient1_tumor_cells_only_fig.svg", width=7, height=5)


```






